String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-27 2023-1-19/sections/cefa.tex
cefa
CEFA

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on.

%\subsection{Cost-Enriched Finite Automaton} \label{subsec:cefa}
The definition of CEFA below is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions on register values
%specified by LIA formulas
%, where the free variables are the registers, 
are attached to final states. 
%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, q_0, F, \theta)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_k\}$ is a finite set of registers, 
    \item $Q, \Sigma, q_0, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma_\varepsilon \times Q \times \Int^R$ is a transition relation, where $\Int^R$ denotes the updates on the values of registers, 
%  set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\alpha: F \rightarrow \Phi(R)$ attaches to each finite state an accepting condition that is specified by an LIA formula.
%    a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
\end{definition}

@@

\begin{example} \label{eg:1}
  Fig.\ref{fig:repetition} illustrates the CEFA recognizing $(ab)\{1,100\}$. The register vector is $(r_1)$, and the linear arithmetic constraint on $q_2$ is $1\leq r_1\leq 100$. Intuitively, the transition $q_0 \xrightarrow[(1)]{a} q_1$ accept the char $a$ and increase 1 on the value of $r_1$. The transition $q_1 \xrightarrow[(0)]{b} q_2$ accepts the char $b$ and does not change the value. The transition $q_2 \xrightarrow[(0)]{\epsilon} q_2$ is a nondeterministic choice back to initial state $q_0$. Because of the linear arithmetic constraint $1\leq r_1\leq 100$, the value of $r_1$ in the accepting run must be $[1,100]$. $\mathcal{V}(r_1)$ equal to the occurrence number of the transition $q_0 \xrightarrow[(1)]{a} q_1$, so that the accepted string is $ab\cdots ab$ in which $a$ appears at least once and at most 100 times. That is, the language of the automaton is $(ab)\{1,100\}$.
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm,
          accepting text=$\scriptstyle 1\leq r_1\leq 100$, accepting/.style=accepting by arrow}
      ]
      \node[state,initial]            (q_0)                      {$q_0$};
      \node[state]                    (q_1) [right=of q_0]       {$q_1$};
      \node[state,accepting]          (q_2) [right=of q_1]       {$q_2$};

>>>>

In this section, we define the model of cost-enriched finite state automata (CEFA), which was introduced in \cite{atva2020} and will be used to solve the satisfiability problem of RECL later on. 
%
Intuitively, CEFA add write-only cost registers to finite state automata. Here by ``write-only'', we mean that the cost registers can only be written/updated, but cannot be read, that is, cannot be used to guard the transitions.

%
%In~\cite{atva2020}, the cost function is defined as a function $\eta: \Sigma \rightarrow \mathbb{N}$. In this paper, we define the cost function as an integer vector whose elements are the incremental value of registers. Furthermore, we add a new linear integer arithmetic constraint $\theta$ to each final state of the CEFA, which restricts the value of registers. Two types of definitions have the same expressive ability on the $SAT_{CL}$ problem.
\begin{definition}[Cost-Enriched Finite Automaton]
  A cost-enriched finite automaton $\aut$ is a tuple $(R, Q, \Sigma, \delta, q_0, F, \theta)$ where
  \begin{itemize}
  \item $R = \{r_1, \cdots, r_k\}$ is a finite set of registers, 
    \item $Q, \Sigma, q_0, F$ are as in the definition of NFA,
%    \item $R = (r_1\cdots r_n)$ is a vector of mutually distinct cost registers,
    \item $\delta \subseteq Q \times \Sigma_\varepsilon \times Q \times \Int^R$ is a transition relation, where $\Int^R$ denotes the updates on the values of registers, 
%  set whose elements are tuples $(q, c, q', \myvec{v})$ where $q, q'$ are states of $Q$, $c$ is a letter in alphabet $\Sigma\cup\{\epsilon\}$ and $\myvec{v}$ is the cost update function for registers, which is an integer vector whose $i$-th element is the incremental value of register $r_i$.  $(q, a, q', \myvec{v})$ is written as $q\xrightarrow[\myvec{v}]{a} q'$ for readability.
    \item $\alpha: F \rightarrow \Phi(R)$ attaches to each finite state an accepting condition that is specified by an LIA formula.
%    a linear integer arithmetic constraint function on final states. $\theta$ is called \emph{accepting condition}.
  \end{itemize}
\end{definition}
The definition of CEFA above is slightly different from the definition in~\cite{atva2020} in the sense that accepting conditions on register values
are attached to final states.

@@

\zhilin{this example should be modified to match the running example}
\begin{example} \label{eg:1}
  Fig.\ref{fig:repetition} illustrates the CEFA recognizing $(ab)\{1,100\}$. The register vector is $(r_1)$, and the linear arithmetic constraint on $q_2$ is $1\leq r_1\leq 100$. Intuitively, the transition $q_0 \xrightarrow[(1)]{a} q_1$ accept the char $a$ and increase 1 on the value of $r_1$. The transition $q_1 \xrightarrow[(0)]{b} q_2$ accepts the char $b$ and does not change the value. The transition $q_2 \xrightarrow[(0)]{\epsilon} q_2$ is a nondeterministic choice back to initial state $q_0$. Because of the linear arithmetic constraint $1\leq r_1\leq 100$, the value of $r_1$ in the accepting run must be $[1,100]$. $\mathcal{V}(r_1)$ equal to the occurrence number of the transition $q_0 \xrightarrow[(1)]{a} q_1$, so that the accepted string is $ab\cdots ab$ in which $a$ appears at least once and at most 100 times. That is, the language of the automaton is $(ab)\{1,100\}$.
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[
      shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
      initial text=, every state/.style={minimum size = 0.001cm,
          accepting text=$\scriptstyle 1\leq r_1\leq 100$, accepting/.style=accepting by arrow}
      ]
      \node[state,initial]            (q_0)                      {$q_0$};
      \node[state]                    (q_1) [right=of q_0]       {$q_1$};
      \node[state,accepting]          (q_2) [right=of q_1]       {$q_2$};
